---
title: "Gov 50 Final Project"
author: "Zoree Jones"
description: "Interesting analysis on intractable challenges."
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 1: Project Thoughts

I am interested in exploring data related to the following topics: (1) Political discourse on social media (particularly as it relates to bias and extremism; (2) international development (e.g., progress towards UN SDGs, intranational inequality); (3) international flows of money (i.e. microfinance, remittances); and (4) legislative productivity. 

## Milestone 2: Project Proposal

Does social media use lead to hostility towards presidential candidates? In this study, I plan to examine the extent to which extended time on social media is related to feelings of anger towards Democratic and Republican presidential candidates. Given the amount of polarizing content that can be found on social media sites like Facebook and Twitter, I predict that more self-reported time on social media will be positively associated with self-reported hostility towards presidential candidates. To examine this relationship, I plan to use data from the 2016 American National Election Survey, which includes 4,271 interviews conducted prior to the 2016 presidential election in English and Spanish. The explanatory variable of interest is the number of days in the week the respondent typically used social media to learn about the presidential election. It is represented as a value from 0 to 7 (days). My outcome variables are described as the respondent's "affect" towards the Republican or Democratic presidential candidate -- specifically whether the frequency with which they reported feeling anger towards the candidate. The variable is coded with the following values: -8: Don't know, 1: Never, 2. Some of the time, 3. About half of the time, 4. Most of the time, and 5. Always. If I observe a positive correlation coefficient between the number of days spent on social media and the degree of anger expressed towards candidates, then I will have support for my hypothesis. However, if I observe that social media usage is either negatively related or unrelated to anger towards candidate, then I will lack sufficient evidence to support my hypothesis.

## Milestone 3: Loading Data + Data Visualization

```{r}
# Load data
library(tidyverse)
library(readr)
anes_2016 <- read_csv("~/Downloads/anes-2016.csv")

# Select relevant variables
anes_subset <- anes_2016 |>
  select(V161116, V161121, V161495)

# Rename variables 
anes_subset <- anes_subset |>
  rename(Dem_Anger = V161116,
         Rep_Anger = V161121,
         Social_Media_Days = V161495)

anes_subset <- anes_subset |>
  mutate(Dem_Anger_Value = case_when(Dem_Anger == 1 ~ "Never",
                                     Dem_Anger == 2 ~ "Some of the time",
                                     Dem_Anger == 3 ~ "About half the time",
                                     Dem_Anger == 4 ~ "Most of the time",
                                     Dem_Anger == 5 ~ "Always") ) |>
  mutate(Rep_Anger_Value = case_when(Rep_Anger == 1 ~ "Never",
                                     Rep_Anger == 2 ~ "Some of the time",
                                     Rep_Anger == 3 ~ "About half the time",
                                     Rep_Anger == 4 ~ "Most of the time",
                                     Rep_Anger == 5 ~ "Always") )

# Turn variables into factors 
anes_subset <- anes_subset |>
  dplyr::mutate(Dem_Anger_Value = factor(Dem_Anger_Value, 
                                    levels = c("Never", "Some of the time", "About half the time",
                                               "Most of the time", "Always"))) |>
   dplyr::mutate(Rep_Anger_Value = factor(Rep_Anger_Value, 
                                    levels = c("Never", "Some of the time", "About half the time",
                                              "Most of the time", "Always"))) 
# Barplot: anger towards Democrats 
dem_anger_plot <- anes_subset |>
  drop_na() |>
  ggplot(mapping = aes(x = Dem_Anger_Value)) + geom_bar(stat = "count", fill = "dodgerblue3") + 
  labs(title = "Respondent Self-Reported Anger Towards 2016 Democratic Presidential Candidate",
       x = "Frequency of feelings of anger towards Democratic presidential candidate",
       y = "Count",
       caption = "Source: American National Election Studies") 
dem_anger_plot

# Barplot: anger towards Republicans  
rep_anger_plot <- anes_subset |>
  drop_na() |>
  ggplot(mapping = aes(x = Rep_Anger_Value)) + geom_bar(stat = "count", fill = "firebrick1") +
  labs(title = "Respondent Self-Reported Anger Towards 2016 Republican Presidential Candidate",
       x = "Frequency of feelings of anger towards Republican presidential candidate",
       y = "Count",
       caption = "Source: American National Election Studies")
rep_anger_plot
```

## Milestone 4: Add Results From One Analysis
```{r}
# Calculate correlation coefficient
cor(anes_subset$Dem_Anger, anes_subset$Social_Media_Days, use = "pairwise")
cor(anes_subset$Rep_Anger, anes_subset$Social_Media_Days, use = "pairwise") 

# Create a linear regression
demfit <- lm(Dem_Anger ~ Social_Media_Days, data = anes_subset)
demfit
repfit <- lm(Rep_Anger ~ Social_Media_Days, data = anes_subset)
repfit

# Table
rep_sm_ate <- anes_subset |> 
  drop_na(Rep_Anger_Value, Social_Media_Days) |>
  mutate(sm_yn = if_else(Social_Media_Days < 1, "No", "Yes")) |>
  group_by(sm_yn) |>
  summarize(mean = mean(Rep_Anger)) |>
  pivot_wider(names_from = sm_yn,
              values_from = mean) |>
  mutate(ATE = `Yes` - `No`) 
  knitr::kable(rep_sm_ate)
rep_sm_ate

dem_sm_ate <- anes_subset |> 
  drop_na(Dem_Anger, Social_Media_Days) |>
  mutate(sm_yn = if_else(Social_Media_Days < 1, "No", "Yes")) |>
  group_by(sm_yn) |>
  summarize(mean = mean(Dem_Anger)) |>
  pivot_wider(names_from = sm_yn,
              values_from = mean) |>
  mutate(ATE = `Yes` - `No`) 
  knitr::kable(dem_sm_ate)
dem_sm_ate
  
```

For exploratory purposes, I began by calculating the correlation coefficients for social media use and feelings towards Republican and Democratic candidates. I discovered a weak positive correlation between social media use and reported anger. For the Democratic candidate, the correlation coefficient is 0.034, while for the Republican candidate, the correlation coefficient is 0.027. However, I did not think that this gave me enough insight into the nature of the linear relationship, so I decided to produce a linear regression. 

I found that the intercept for the Democratic candidate is 2.59, meaning that when the Social_Media_Days variable is 0 (an individual does not get any election news from social media), the predicted self-reported anger value is between feeling angry "some of the time" and "about half of the time." For each unit increase in the number of days on social media, we can predict a small increase in the dependent variable of 0.016. The Republican candidate intercept was slightly higher, indicating that there may have been a slightly higher baseline level of anger towards them compared to the Democratic candidate. However, the effect of social media use on anger per unit increase was smaller, at just 0.012.

Finally, I produced two ATE tables to identify whether there was a notable treatment effect of using social media at all. I turned the Social_Media_Days variable into a binary variable of either no use (0 days) or some use (1-7 days). Here, I again saw small magnitudes, with the ATE for the Republican candidate being 0.068 and the ATE for the Democratic candidate being 0.083. However, the slightly larger ATE for the Democratic candidate is unsurprising given the larger step increase predicted by the linear model.

Overall, these findings indicate that social media use as measured in this study is not a particularly strong predictor of increased anger towards presidential candidates of either party.

