---
title: "Gov 50 Final Project"
author: "Zoree Jones"
description: "Thanks for a great class! :)"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction

Understanding political polarization and its causes is of central interest to many political scientists. Defined as …Within the last few decades, polarization has risen sharply among Americans – a Pew Research Center poll conducted in the run-up to the 2020 presidential election reported that 90 percent of both Trump and Biden supporters believed that if the other candidate won, it would cause “lasting harm” to the country.  [Sentence about democratic threat of polarization]. 

In an aim to identify what distinguishes this period of American polarization from past historical moments, some observers have pointed to social media platforms as a key factor. One meta-analysis of the relevant social science literature conducted by the Brookings Institution concluded that while “platforms like Facebook, YouTube, and Twitter likely are not the root causes of political polarization, they do exacerbate it.” 

In light of the increased attention on social media platforms as a contributor to polarization, I was interested in examining whether it was possible to draw a direct association between social media and feelings of animus. Using data from the 2016 American National Election Studies time series survey, I sought to answer the following research question: Is the use of social media to learn about presidential elections positively associated with self-reported feelings of hostility towards presidential candidates?

Given the amount of polarizing content that can be found on social media sites like Facebook and Twitter, I predicted that more self-reported use of social media to learn about presidential elections would be positively associated with self-reported hostility towards presidential candidates.


## Milestone 2: Project Proposal

Does social media use lead to hostility towards presidential candidates? In this study, I plan to examine the extent to which extended time on social media is related to feelings of anger towards Democratic and Republican presidential candidates. Given the amount of polarizing content that can be found on social media sites like Facebook and Twitter, I predict that more self-reported time on social media will be positively associated with self-reported hostility towards presidential candidates. To examine this relationship, I plan to use data from the 2016 American National Election Survey, which includes 4,271 interviews conducted prior to the 2016 presidential election in English and Spanish. The explanatory variable of interest is the number of days in the week the respondent typically used social media to learn about the presidential election. It is represented as a value from 0 to 7 (days). My outcome variables are described as the respondent's "affect" towards the Republican or Democratic presidential candidate -- specifically whether the frequency with which they reported feeling anger towards the candidate. The variable is coded with the following values: -8: Don't know, 1: Never, 2. Some of the time, 3. About half of the time, 4. Most of the time, and 5. Always. If I observe a positive correlation coefficient between the number of days spent on social media and the degree of anger expressed towards candidates, then I will have support for my hypothesis. However, if I observe that social media usage is either negatively related or unrelated to anger towards candidate, then I will lack sufficient evidence to support my hypothesis.

## Data 

The data for this analysis came from the 2016 American National Election Studies time series survey. The survey consisted of a pre-election questionnaire administered between September 7 and November 7, 2016 and a post-election questionnaire administered between November 9 and January 8, 2017. The survey reached a final sample size of 4,271. The target population for the study was the 224.1 million U.S. citizens who are age 18 or older.

The key dependent variables for this analysis are defined as the respondent’s “affect” towards the Republican or Democratic presidential candidate – specifically whether the frequency with which they reported feeling anger towards the candidate. The variables are coded with the following values: -8: Don’t know, 1: Never, 2. Some of the time, 3. About half of the time, 4. Most of the time, and 5. Always. The independent/explanatory variable of interest is the number of days in the week the respondent typically used social media to learn about the presidential election. It is represented as a value from 0 to 7 (days).

The following bar charts represent the distribution of the dependent variable (how often they reported feeling angry towards the Democratic and Republican presidential candidate).

The remainder of my analysis proceeded in the following way: 
(1) First, I calculated the correlation coefficient to get a sense of the direction of the relationship. 
(2) Next, I produced a series of linear regressions. The first two regressions captured the predicted change in the dependent variable given a one unit increase in the number of days on social media.The second two regressions captured the predicted change in the dependent variable based on any use of social media using the modified binary sm_yn variable, which captured whether or not a respondent used social media to learn about the presidential election at all.


```{r}
# Load data
library(tidyverse)
library(readr)
anes_2016 <- read_csv("~/Downloads/anes-2016.csv")

# Select relevant variables
anes_subset <- anes_2016 |>
  select(V161116, V161121, V161495)

# Rename variables 
anes_subset <- anes_subset |>
  rename(Dem_Anger = V161116,
         Rep_Anger = V161121,
         Social_Media_Days = V161495)

anes_subset <- anes_subset |>
  mutate(Dem_Anger_Value = case_when(Dem_Anger == 1 ~ "Never",
                                     Dem_Anger == 2 ~ "Some of the time",
                                     Dem_Anger == 3 ~ "About half the time",
                                     Dem_Anger == 4 ~ "Most of the time",
                                     Dem_Anger == 5 ~ "Always") ) |>
  mutate(Rep_Anger_Value = case_when(Rep_Anger == 1 ~ "Never",
                                     Rep_Anger == 2 ~ "Some of the time",
                                     Rep_Anger == 3 ~ "About half the time",
                                     Rep_Anger == 4 ~ "Most of the time",
                                     Rep_Anger == 5 ~ "Always") )

# Turn variables into factors 
anes_subset <- anes_subset |>
  dplyr::mutate(Dem_Anger_Value = factor(Dem_Anger_Value, 
                                    levels = c("Never", "Some of the time", "About half the time",
                                               "Most of the time", "Always"))) |>
   dplyr::mutate(Rep_Anger_Value = factor(Rep_Anger_Value, 
                                    levels = c("Never", "Some of the time", "About half the time",
                                              "Most of the time", "Always"))) 
# Barplot: anger towards Democrats 
dem_anger_plot <- anes_subset |>
  drop_na() |>
  ggplot(mapping = aes(x = Dem_Anger_Value)) + geom_bar(stat = "count", fill = "dodgerblue3") + 
  labs(title = "Respondent Self-Reported Anger Towards 2016 Democratic Presidential Candidate",
       x = "Frequency of feelings of anger towards Democratic presidential candidate",
       y = "Count",
       caption = "Source: American National Election Studies") 
dem_anger_plot

# Barplot: anger towards Republicans  
rep_anger_plot <- anes_subset |>
  drop_na() |>
  ggplot(mapping = aes(x = Rep_Anger_Value)) + geom_bar(stat = "count", fill = "firebrick1") +
  labs(title = "Respondent Self-Reported Anger Towards 2016 Republican Presidential Candidate",
       x = "Frequency of feelings of anger towards Republican presidential candidate",
       y = "Count",
       caption = "Source: American National Election Studies")
rep_anger_plot
```

## Results 
```{r}

# Add binary variable for social media use 
anes_subset <- anes_subset |>
  mutate(sm_yn = if_else(Social_Media_Days < 1, "No", "Yes"))
         
# Calculate correlation coefficient
cor(anes_subset$Dem_Anger, anes_subset$Social_Media_Days, use = "pairwise")
cor(anes_subset$Rep_Anger, anes_subset$Social_Media_Days, use = "pairwise") 

# Create a linear regression (social media as a categorical variable)
demfit <- lm(Dem_Anger ~ Social_Media_Days, data = anes_subset)
demfit
modelsummary::modelsummary(demfit,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

repfit <- lm(Rep_Anger ~ Social_Media_Days, data = anes_subset)
modelsummary::modelsummary(repfit,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

# Create a linear regression (social media as a binary variable)
demfit2 <- lm(Dem_Anger ~ sm_yn, data = anes_subset)
demfit2

repfit2 <- lm(Rep_Anger ~ sm_yn, data = anes_subset)
repfit2

# Calculate R^2 for initial models 
demfitsum <- summary(demfit)
demfitsum$r.squared

repfitsum <- summary(repfit)
repfitsum$r.squared

# Create additional linear regressions 
newfit <- lm(V161116 ~ V161495 + V161081 + V161162 + V161270 + V161169, data = anes_2016)
var_labels <- c(
  "V161116" = "Rep_Anger_Value",
  "V161495" = "Social_Media_Days",
  "V161081" = "Right vs. wrong track",
  "V161162" = "Dem. candidate is honest",
  "V161270" = "Highest education level",
  "V161169" = "Dem. candidate is even tempered"
)

modelsummary::modelsummary(newfit,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

```

For exploratory purposes, I began by calculating the correlation coefficients for social media use and feelings towards Republican and Democratic candidates. I discovered a weak positive correlation between social media use and reported anger. For the Democratic candidate, the correlation coefficient is 0.034, while for the Republican candidate, the correlation coefficient is 0.027. However, I did not think that this gave me enough insight into the nature of the linear relationship, so I decided to produce a linear regression. 

I found that the intercept for the Democratic candidate is 2.59, meaning that when the Social_Media_Days variable is 0 (an individual does not get any election news from social media), the predicted self-reported anger value is between feeling angry “some of the time” and “about half of the time.” For each unit increase in the number of days on social media, we can predict a small increase in the dependent variable of 0.016. The Republican candidate intercept was slightly higher at 2.91, indicating that there may have been a slightly higher baseline level of anger towards them compared to the Democratic candidate. However, the effect of social media use on anger per unit increase was smaller, at just 0.013.

To identify whether there was a notable treatment effect of using social media at all, I turned the Social_Media_Days variable into a binary variable called sm_yn of either no use (0 days) or some use (1-7 days). After doing so, I found that the intercept for the Democratic candidate was also 2.59. Use of social media predicted a small (0.08) point increase in anger value. The Republican candidate intercept was 2.91, and the effect of social media use on anger was 0.07.

Next, I examined the explanatory power of the regressions. The first regression "demfit" has an R^2 of 0.001. The second regression "repfit" also has an R^2 of 0.001. Both values are extremely small, indicating that the vast majority of the variance observed in the dependent variable (anger) cannot be explained by variance in amount of time on news-related social media.

Finally, in an attempt to develop a model with better explanatory power, I selected the following additional variables: (1) Are things in the country on the right track? (binary variable - right vs. wrong track); (2) Is the democratic candidate honest? (1-5 scale); (3) Highest level of education; (4) Is the democratic candidate even-tempered? (1-5 scale). This model appears to be better equipped to explain variation in the dependent variable, as it has an R^2 of 0.495.

Overall, these findings indicate that social media use as measured in this study is not a particularly strong predictor of increased anger towards presidential candidates of either party. 

